Social Attention for Autonomous Decision-Making in Dense Traffic

### 2019-09-14 paper/main.tex
Last changes

In order to apply a reinforcement learning algorithm such as DQN to an autonomous driving problem, a state space $S$ must first be chosen, that is, a representation of the scene. When social interactions are relevant to the decision, the state should at least compose a description of every nearby vehicle. A vehicle driving on a road can be described in the most general way by it's continuous position, heading and velocity. Then, the joint state of a road traffic with one ego-vehicle denoted $s_0$ and $N$ other vehicles can be described by the (unordered) set of individual vehicle states:

\begin{equation}
s = \left\lbrace s_i \right\rbrace_{i \in [0, N]}\qquad
\text{where}\qquad
s_i = \begin{bmatrix}
x_i & y_i & v^x_i & v^y_i & \cos\psi_i & \sin \psi_i
\end{bmatrix}^T
\label{eq:coordinates}
\end{equation}

@@

This encoding is efficient in the sense that it uses the smallest quantity of information necessary to represent the scene. However, it lacks two important properties. First, its size varies with the number of vehicles which can be problematic for the sake of function approximation which often expects constant-sized inputs. Second, we expect a driving policy $\pi$ to be \emph{permutation invariant}, i.e. not to be dependent on the order in which other traffic participants are listed. Ideally, this property should be enforced and not approximated by relying on the coverage of the $N!$ possible permutations $\tau$ of any given traffic state. Formally, we require that:
\begin{equation*}
\pi(\cdot|(s_0, s_1,\dotsc,s_N)) = \pi(\cdot|(s_0, s_{\tau(1)},\dotsc,s_{\tau(N)})) \quad\quad\quad \forall\tau \in \mathfrak{S}_N
\end{equation*}

@@

This permutation invariance property can also be implemented within the policy architecture. A general technique to achieve this is to treat each entity similarly in the early stages -- e.g. through weight sharing -- before reducing them with a projection operator that is itself invariant to permutations, for instance a max-pooling as in \citep{Chen2017} or an average as in \citep{Qi2016}. A particular instantiation of this idea is attention mechanisms.

@@

In this context, the agent must drive a vehicle by controlling its acceleration chosen from a finite set of actions $A = \{\texttt{SLOWER}, \texttt{NO-OP}, \texttt{FASTER}\}$. The lateral control is performed automatically by a low-level controller, such that the problem complexity is focused on the high-level interactions with other vehicles, namely the decision to either yield or take way. The agent is rewarded by $1$ when it drives at maximum velocity, $0$ otherwise, and by $-5$ when a collision occurs.

@@

We plot in \autoref{fig:results} the evolution of the total reward, episode length and average velocity during training, over 4000 episodes and repeated across 120 random seeds.
The \MLPL agent learns to accelerate to earn short-term rewards, as shown by its high average velocity, but fails to exploit the information of other vehicles and crashes often, leading to shorter episodes. In short, we obtain a risky and blind policy that is the worst performing.
Conversely, the \MLPG architecture benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions, as we can see from its higher episode length. However, it only proceeds when the intersection has been fully cleared, as reflected by its low average velocity. This results in an overly cautious policy -- a common trait colloquially known as the \emph{freezing robot problem} \citep{Trautman2010} -- with a slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection (high episode length), but also when it can exploit the gaps in the traffic and take way to vehicles that are far or slow enough (higher average velocity than \MLPG). This translates as a significant performance improvement, and the overall resulting behaviour is qualitatively more nuanced and human-like.

@@

In any given state, the attention matrix can be visualized in the following way: we connect the ego-vehicle to every vehicle by a line of width proportional to the corresponding attention weight. Since the architecture can contain several Ego-Attention heads, we use different colours to distinguish them. In our experiments, two attention heads were used and will be represented in green and blue. We observe in \autoref{fig:heads} that they specialized to focus on different areas: the green head is only watching the vehicles coming from the left, while the blue head restricts itself to vehicles in the front and right directions. However, we notice that both heads exhibit a common behaviour: they focus their attention on incoming vehicles that are likely to collide with the ego-vehicle, depending on their current position, heading, velocity, and ignore those that are too far or in a conflict-less situation. In particular, the attention tends to increase when vehicles get closer, as shown in \autoref{fig:distance}. It can also be very sensitive to small variations in the traffic state, as reflected in \autoref{fig:sensitivity}. A full episode showcasing interactions with several vehicles is shown in \autoref{fig:episode}.

@@

The priority rules are not enforced through rewards but interactions: based on the defined road priorities, some vehicles will take way to the ego-vehicle while others will not. By changing which is a priority road, we can influence the rules of interactions which affects the learnt behaviour. In \autoref{fig:priority}, we compare two policies placed in the exact same initial state and observe how their decisions are affected by their model of how incoming vehicles interact with them. This difference showcases the ability of our proposed architecture to discover and exploit such interaction patterns.

>>>>

In order to apply a reinforcement learning algorithm such as DQN to an autonomous driving problem, a state space $S$ must first be chosen, that is, a representation of the scene. When social interactions are relevant to the decision, the state should at least contain a description of every nearby vehicle. A vehicle driving on a road can be described in the most general way by it's continuous position, heading and velocity. Then, the joint state of a road traffic with one ego-vehicle denoted $s_0$ and $N$ other vehicles can be described by a list of individual vehicle states:

\begin{equation}
s = \left( s_i \right)_{i \in [0, N]}\qquad
\text{where}\qquad
s_i = \begin{bmatrix}
x_i & y_i & v^x_i & v^y_i & \cos\psi_i & \sin \psi_i
\end{bmatrix}^T
\label{eq:coordinates}
\end{equation}

@@

This encoding is efficient in the sense that it uses the smallest quantity of information necessary to represent the scene. However, it lacks two important properties. First, its size varies with the number of vehicles which can be problematic for the sake of function approximation which often expects constant-sized inputs. Second, we expect a driving policy $\pi$ to be \emph{permutation invariant}, i.e. not to be dependent on the order in which other traffic participants are listed. Ideally, this property should be enforced and not approximated by relying on the coverage of the $N!$ possible permutations $\tau$ of any given traffic state. Formally, we require that:
\begin{equation}
\label{eq:permutation}
\pi(\cdot|(s_0, s_1,\dotsc,s_N)) = \pi(\cdot|(s_0, s_{\tau(1)},\dotsc,s_{\tau(N)})) \quad\quad\quad \forall\tau \in \mathfrak{S}_N
\end{equation}

@@

This permutation invariance property \eqref{eq:permutation} can also be implemented within the architecture of the policy $\pi$. A general technique to achieve this is to treat each entity similarly in the early stages -- e.g. through weight sharing -- before reducing them with a projection operator that is itself invariant to permutations, for instance a max-pooling as in \citep{Chen2017} or an average as in \citep{Qi2016}. A particular instance of this idea is attention mechanisms.

@@

In this context, the agent must drive a vehicle by controlling its acceleration chosen from a finite set of actions $A = \{\texttt{SLOWER}, \texttt{NO-OP}, \texttt{FASTER}\}$. The lateral control is performed automatically by a low-level controller, such that the problem complexity is focused on the high-level interactions with other vehicles, namely the decision to either give or take way. The agent is rewarded by $1$ when it drives at maximum velocity, $0$ otherwise, and by $-5$ when a collision occurs.

@@

We plot in \autoref{fig:results} the evolution of the total reward, episode length and average velocity during training, over 4000 episodes and repeated across 120 random seeds.
The \MLPL agent learns to accelerate to earn short-term rewards, as shown by its high average velocity, but fails to exploit the information of other vehicles and crashes often, leading to short episodes. We obtain a risky and blind policy that is the worst performing.
Conversely, the \MLPG architecture benefits from its invariance to permutations and manages to learn to brake upon arrival at the intersection to avoid collisions, as we can see from its higher episode length. However, it only proceeds when the intersection has been fully cleared, as reflected by its low average velocity. This results in an overly cautious policy -- a common trait colloquially known as the \emph{freezing robot problem} \citep{Trautman2010} -- with a slight increase in performance.
In stark contrast, the \EgoAtt policy quickly learns both when it must slow down at the intersection (see the high episode length), but also when it can exploit the gaps in the traffic and take way to vehicles that are far or slow enough (see the higher average velocity than \MLPG). This translates as a significant performance improvement, and the overall resulting behaviour is qualitatively more nuanced and human-like.

@@

In any given state, the attention matrix can be visualized in the following way: we connect the ego-vehicle to every vehicle by a line of width proportional to the corresponding attention weight. Since the architecture can contain several ego-attention heads, we use different colours to distinguish them. In our experiments, two attention heads were used and will be represented in green and blue. We observe in \autoref{fig:heads} that they specialized to focus on different areas: the green head is only watching the vehicles coming from the left, while the blue head restricts itself to vehicles in the front and right directions. However, we notice that both heads exhibit a common behaviour: they direct their attention to incoming vehicles that are likely to collide with the ego-vehicle, depending on their current position, heading, velocity, and ignore those that are too far or in a conflict-less situation. In particular, the attention tends to increase when vehicles get closer, as shown in \autoref{fig:distance}. It can also be very sensitive to small variations in the traffic state, as reflected in \autoref{fig:sensitivity}. A full episode showcasing interactions with several vehicles is shown in \autoref{fig:episode}.

@@

The agent decisions regarding right of way are not enforced through rewards but interactions: based on the defined road priorities, some vehicles will take way to the ego-vehicle while others will not. By changing which is a priority road, we can influence the rules of interactions which affects the learnt behaviour. In \autoref{fig:priority}, we compare two policies placed in the exact same initial state and observe how their decisions are affected by their internal model of how incoming vehicles interact with them. This difference showcases the ability of our proposed architecture to discover and exploit such interaction patterns.
